---
title: "gapminder"
author: "Ingrid"
date: "March 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Notebook on how to do all of this "http://ohi-science.org/data-science-training/dplyr.html"
# Go to help-cheatsheets-data transformation with dplyr
#load packages
#Shortcuts Ctrl and Enter (run that line of code); Alt and - (<-), put cursor in the function and press f1-gives you thehelp function
```{r}
library(tidyverse)
```
# the red crosses mean that the function (e.g. filter) in diplyr will mask the default filter. Have to use "stats::filter"" to use the default filter instead
#gapminder dataset
```{r}
gapminder <-  read.csv("https://raw.githubusercontent.com/jules32/2017-11-30-MBARI/gh-pages/data/gapminder.csv")
```
```{r}
head(gapminder, 10)
tail(gapminder)
```
```{r}
str(gapminder)
```

```{r}
dim(gapminder)
ncol(gapminder)
nrow(gapminder)
```
```{r}
#list unique and then select a column from the dataset to list e.g the number of countries in dataset, with $ sign if the beginning of string is the same it might miss it out of the final list
unique(gapminder$country)
#or unique(gapminder[,"continent"]), which is more specific
```
```{r}
summary(gapminder)
```
```{r}
#install.packages("skimr")
library(skimr)
skim(gapminder)
```

```{r}
filter(gapminder, country == "New Zealand")
```

```{r}
filter(gapminder,  lifeExp <30)
```

```{r}
filter(gapminder, country == "Afghanistan", year == max(year))
```
#starting to play with dplyr
```{r}
dplyr::select(gapminder, country, year, pop)
#selec is another way to filter, but gives all the results for that country

```

```{r}
dplyr::select(gapminder, -continent)
#can also see everything, excluding select columns
```
```{r}
gapminder_no_cont <- dplyr::select(gapminder, -continent)
dplyr::filter(gapminder_no_cont, year ==2007)
```

```{r}
# %>% = a pipe (take the result from the left and put it into the function on the left)-same as above without having to make a new object, and don't have to specifify the worksheet "gapminder"
select(gapminder, -continent) %>% filter(year == 2007)

```

```{r}
gapminder %>%
  mutate(gdp = gdpPercap * pop)
```

```{r}
gapminder %>%
  mutate(gdp = gdpPercap * pop) %>%
  arrange (desc(gdp))
```
```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp = gdpPercap * pop, max_gdp = max(gdp)) %>%
  ungroup() 
#need to ungroup after to get back to the full dataset instead of the grouped data
```
```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp = pop *gdpPercap) %>%
  summarize(max_gdp = max(gdp))
#to collapse into one summary column use summarize
```
```{r}
#Joining tables
#dplyr::left_join: join all the tables relative to the first table (on the left)- opposite for right
#dplyr::inner_join: join common set of rows (so removes rows which don't match all the way with NA's)
#dplyr::full_join: includes all columns and adds NA
co2 <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
```
```{r}
head(co2)
summary(co2)
```
```{r}
gap_2007 <- gapminder %>%
  filter(year == 2007)
```

```{r}
#just to make sure it got every country
length(unique(gapminder$country))
```
```{r}
left_join(gap_2007, co2, by = "country")
```
```{r}
right_join(gap_2007, co2, by = "country")
#anti-join or set diff are ways of finding where they don't match
```

##Tidyr
```{r}
gap_wide <- read.csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv")

```




```{r}
gap_better <- gap_wide %>%
  gather(observation, observ_value, -continent, -country) %>%
  separate(observation, into = c("obs_type", "year"), sep = "_", convert = TRUE)
#convert default is FALSE (changed year from a character to an integer)
gap_better
```
```{r}
unique(gap_better$obs_type)
```
```{r}
gap_better %>%
  spread(obs_type, observ_value)
#everthing was gathered into one column, and now those 3 observation types have been seperated into 3 columns
```

